---
title: "1.0-bus-service-ridership"
author: "Julian Beckert"
format: html
editor: visual
---

## Libraries and datasets

```{r}
library(tidyverse)
# RideOn GTFS feed data
rideon_stop_times <- read_delim("rideon_stop_times.txt")
rideon_trips <- read_delim("rideon_trips.txt")
rideon_routes <- read_delim("rideon_routes.txt")
rideon_calendar <- read_delim("rideon_calendar.txt")

# Ridership
rideon_ridership_feb26 <- read_csv("Feb2026RideOnAvgDailyRidership.csv")
wmata_ridership_feb26 <- read_csv("Feb26WMATAAvgDailyRidership.csv")

# WMATA cleaned GTFS feed data 
# NOTE: This file is NOT included in the GitHub repository
wmata_all_calendar_trips <- read_csv("wmata_all_calendar_trips.csv")

```

```{r}
rideon_calendar
```

**Note:**

`service_id` of:\
**1** means service on Monday-Friday\
**2** means Saturday\
**3** means Sunday\

### RideOn trips per route/day summary

```{r}
# join different GTFS files together
rd_calendar_trips1 <- left_join(rideon_calendar, rideon_trips, by = "service_id")
rd_calendar_trips <- left_join(rd_calendar_trips1, rideon_stop_times, by = "trip_id")

rd_calendar_stops_trips <- left_join(rd_calendar_trips, rideon_routes)

rd_weekday_calendar_trips <- rd_calendar_stops_trips %>% 
  filter(service_id == 1) %>% # service_id == 1 means runs monday-friday
  group_by(route_short_name) %>% 
  summarise(
    wkday_departures = length(unique(trip_id))
  )

rd_sat_calendar_trips <- rd_calendar_stops_trips %>% 
  filter(service_id == 2) %>% # service_id == 2 means runs saturday
  group_by(route_short_name) %>% 
  summarise(
    sat_departures = length(unique(trip_id))
  )

rd_sun_calendar_trips <- rd_calendar_stops_trips %>% 
  filter(service_id == 3) %>% ## service_id == 3 means runs sunday
  group_by(route_short_name) %>% 
  summarise(
    sun_departures = length(unique(trip_id))
  )

# join into one to show number of trips by day of week
rd_all_calendar_trips <- left_join(rd_weekday_calendar_trips, rd_sat_calendar_trips, by = "route_short_name")
rd_all_calendar_trips <- left_join(rd_all_calendar_trips, rd_sun_calendar_trips, by = "route_short_name")

rd_all_calendar_trips[is.na(rd_all_calendar_trips)] <- 0 # replace NA with 0 


sum(rd_all_calendar_trips$sat_departures)
sum(rd_all_calendar_trips$sun_departures)
```

## Ridership

Clean and combine RideOn + WMATA Metrobus ridership

```{r}
rideon_ridership_feb26 <- rideon_ridership_feb26 %>% 
  select(-Holiday) %>%  # Metrobus doesn't have this column, so removing
  rename_at("Route",~"route_name")

wmata_ridership_feb26 <- wmata_ridership_feb26 %>% 
  rename_at("Route",~"route_name")
wmata_ridership_feb26[is.na(wmata_ridership_feb26)] <- 0

joined_ridership_feb26 <- bind_rows(rideon_ridership_feb26, wmata_ridership_feb26)

# make the naming more standard
ridership_feb26 <- joined_ridership_feb26 %>% 
  rename_at("Weekday",~"wkday_riders") %>% 
  rename_at("Saturday",~"sat_riders") %>% 
  rename_at("Sunday",~"sun_riders")
```

The RideOn ridership file used "XPINK" and "XLIME" instead of PINK and LIME, so I renamed them in the csv file itself. Likewise, the RideOn ridership data includes three Flex routes (Flex Rockville, Flex Wheaton, and Flex 918) and bus route 301, that are not included in the GTFS files, and will be removed when joining to the GTFS files.

```{r}
wmata_all_calendar_trips <- wmata_all_calendar_trips %>%
  rename_at("route_short_name",~"route_name") %>%
  mutate(agency = "WMATA")
rd_all_calendar_trips <- rd_all_calendar_trips %>%
  rename_at("route_short_name",~"route_name") %>%
  mutate(agency = "MC RideOn")
all_calendar_trips <- bind_rows(rd_all_calendar_trips, wmata_all_calendar_trips)

sum(all_calendar_trips$wkday_departures)
sum(all_calendar_trips$sat_departures)
sum(all_calendar_trips$sun_departures)

buses_calendar_ridership <- left_join(all_calendar_trips, ridership_feb26)

# NOTE: This file IS included in the GitHub repository
# write_csv(buses_calendar_ridership, "metrobus_rideon_all_trips_calendar_schedule.csv")
```
