---
title: "3.0-analysis-access"
author: "Julian Beckert"
format: html
editor: visual
---

## Libraries & datasets

```{r}
library(tidyverse)
bus_data <- read_csv("final_bus_access_dataset_simple.csv")
city_tract_size <- read_csv("censusandcitysize.csv")
all_calendar_trips <- read_csv("mocobuses_all_trips_calendar_schedule.csv") 
bus_tracts_summary <- read_csv("bus_tracts_acs_summary.csv")
```

The aim of this project is to communicate the extent of bus service in Montgomery County and contextualize it with census estimates that indicate need for bus service.

## Statistics: Percentages and totals

```{r}
totaldwellings <- sum(bus_data$totalDwellings)
totalhouseholds_est <- sum(bus_data$estHouseholds)
totalpopulationest <- sum(bus_data$estPopulation)

# I tried getting this with base R, but decided tidyverse would be simpler:
mostUsedBus_tibble <- bus_tracts_summary %>% 
  mutate(most_route = 
           wkdayRiders+satRiders+sunRiders
         ) %>% 
  filter(squareMilesOfTracts > 3) %>%  # this filters out a lot of Metrobus routes that barely enter Montgomery County at all
  arrange(desc(most_route))
mostUsedBus <- as.character(mostUsedBus_tibble$route_name[which.max(mostUsedBus_tibble$most_route)])

mostUsedRideOnBus_tibble <- bus_tracts_summary %>% 
    mutate(most_route = 
           wkdayRiders+satRiders+sunRiders
         ) %>% 
  filter(agency == "MC RideOn") %>%
  arrange(desc(most_route))
mostUsedRideOnBus <- as.character(mostUsedRideOnBus_tibble$route_name[which.max(mostUsedRideOnBus_tibble$most_route)])

  

cat(paste0(
  "Most census tracts have about ", median(bus_data$totalRoutes), " bus routes crossing through them.\n",
  "Most buses pass through about ", median(bus_tracts_summary$tracts), " census tracts.\n",
  "The most used bus week-round in Montgomery County is ", mostUsedBus, " with ", mostUsedBus_tibble$wkdayDepartures[mostUsedBus_tibble$route_name == mostUsedBus], " departures and an average of ", mostUsedBus_tibble$wkdayRiders[mostUsedBus_tibble$route_name == mostUsedBus], " riders per weekday.\n",
  "The most used RideOn bus week-round in Montgomery County is ", mostUsedRideOnBus, " with ", mostUsedRideOnBus_tibble$wkdayDepartures[mostUsedRideOnBus_tibble$route_name == mostUsedRideOnBus], " departures and an average of ", mostUsedRideOnBus_tibble$wkdayRiders[mostUsedRideOnBus_tibble$route_name == mostUsedRideOnBus], " riders per weekday.\n",
  
  "According to estimates by the American Community Survey,\n", 
  "  • ", round(sum(bus_data$estHouseholds_MorePeopleThanVehicles)/totalhouseholds_est,4)*100, "% of households in Montgomery County have more people than vehicles.\n",
  "  • ", round(sum(bus_data$estHouseholds_1Vehicle_2PlusPeople)/totalhouseholds_est,4)*100, "% of households have one vehicle and more than two people.\n",
  "  • ", round(sum(bus_data$estHouseholds_LessVehiclesThanWorkers)/totalhouseholds_est,4)*100, "% of households have less vehicles than workers.\n",
  "  • ", round(sum(bus_data$estHouseholds_0Vehicle)/totalhouseholds_est,4)*100, "% of households have 0 vehicles altogether.\n",

  
  "When defining 'home' as every single-family residential building and every individual unit in a multi-family residential building,\n",
  "  • ", round(sum(bus_data$quarterm_stop_totalDwellings)/totaldwellings,4)*100, "% of homes are within walking distance (1/4 mile) of a bus stop.\n",
  "  • ", round(sum(bus_data$halfm_stop_totalDwellings)/totaldwellings,4)*100, "% of homes are within 1/2 mile of a bus stop.\n",
  "  • ", sum(bus_data$quarterm_route_BRT_totalDwellings), " homes are within walking distance (1/4 mile) of a proposed or existing Bus Rapid Transit route/line. This is ", round(sum(bus_data$quarterm_route_BRT_totalDwellings)/totaldwellings,4)*100, "% of all homes. \n",
"  • ", sum(bus_data$halfm_route_BRT_totalDwellings), " homes are within 1/2 mile of a proposed or existing BRT route/line. This is ", round(sum(bus_data$halfm_route_BRT_totalDwellings)/totaldwellings,4)*100, "% of all homes. \n",
  
  "\nIt is important to note that being in walking distance of a bus stop is not the same thing as having reliable bus service. This is not something fully measured in the scope of this project, but: \n",
  "  • Approximately ", round(sum(bus_data$estHouseholds[bus_data$is_sunday == 0])/sum(bus_data$estHouseholds),4)*100, "% of households are in a census tract that has no Sunday service, accounting for approximately ", round(sum(bus_data$estPopulation[bus_data$is_sunday == 0])/sum(bus_data$estPopulation),4)*100, "% of Montgomery County's population.\n",
  "  • The total departures per day drops ", 100-(round(sum(all_calendar_trips$sat_departures)/sum(all_calendar_trips$wkday_departures),4)*100), "% from weekdays to Saturdays, going from ", sum(all_calendar_trips$wkday_departures), " to ", sum(all_calendar_trips$sat_departures),  " trips.\n",
  "  • The total departures per day drops ", 100-(round(sum(all_calendar_trips$sun_departures)/sum(all_calendar_trips$wkday_departures),4)*100), "% from weekdays to Sundays, going from ", sum(all_calendar_trips$wkday_departures), " to ", sum(all_calendar_trips$sun_departures),  " trips.\n",
  "  • On Saturdays, most census tracts have around ", median(bus_data$totalSatDepartures), " departures, and ", sum(bus_data$totalSatDepartures < 100), " census tracts have less than 100 departures per day on Saturday, accounting for about ", round(sum(bus_data$estPopulation[bus_data$totalSatDepartures < 100])/totalpopulationest, 4)*100, "% of the population."
))
```

## Tracts of interest

I want to highlight census tracts that indicate that they might benefit the most from improved bus service. To do this, I'm looking mainly at two variables: the proportion of households with less vehicles than workers (`propEstHouseholds_LessVehiclesThanWorkers`) and the proportion of dwellings that are a quarter mile from a bus stop (`propDwell_quarterm_stop`).

```{r}
med_vehicles <- median(bus_data$propEstHouseholds_LessVehiclesThanWorkers)
med_quarterm <- median(bus_data$propDwell_quarterm_stop)

bus_data %>% 
  mutate(has_brt_stop_cat = ifelse(has_brt_stop == 1, "Yes", "No"),
         buses_per_square_mile = totalRoutes/SquareMile) %>% 
  ggplot() +
  # annotate before geom_point so it won't cover the points
  annotate("rect", xmin = med_vehicles, ymin = med_quarterm, xmax = 0.45, ymax = 0.00, fill = "#F8509C", alpha = 0.2) +
  geom_point(aes(
    x = propEstHouseholds_LessVehiclesThanWorkers, y = propDwell_quarterm_stop, 
    color = has_brt_stop_cat,
    size = buses_per_square_mile
    ), alpha = 0.9) +
  
  geom_vline(xintercept = 
               med_vehicles) +
  geom_hline(yintercept = 
               med_quarterm) +
  
  scale_color_manual(values = c("#141414", "#F8509C"), aesthetics = "color") +
  
  labs(
    title = "Census tracts: proportion of homes 1/4 mile from a bus stop vs\nproportion of households with less vehicles than workers",
    x = "Estimated proportion of households with less vehicles than workers",
    y = "Proportion of dwellings 1/4 mile from bus stop",
    color = "Has BRT stop",
    size = "Buses per square mile"
       ) +
  theme_light() +
  theme(
    axis.line = element_line(linewidth = 0.75, color = "#141414"),
    panel.border = element_rect(linewidth = 0.45, color = "#FF93C4", fill = "#F9EFF6"),
    panel.grid.major.y = element_line(color = "#F9EFF6"),
                      # (left, bottom, right, top) 
    legend.position = c(0.75, 0.4, 0.1, 0.4),
    legend.key.size = unit(0.5, 'cm'), #change legend key size
    axis.ticks = element_blank(),
    axis.text.x = element_text(color = "#E7CEE0"),
    axis.text.y = element_text(color = "#E7CEE0"),
    plot.background = element_rect(fill = "#141414", color = "#FF93C4"),
    plot.title = element_text(color = "#fff", size = 13, hjust = 0.5),
    axis.title = element_text(color = "#fff"),
    plot.subtitle = element_text(color = "#E7CEE0")
  )
```

This scatterplot shows a lot of variables at once, but again, the main two most important variables are the proportion of dwellings 1/4 mile from a bus stop and the proportion of households with less vehicles than workers. I am selecting all the census tracts that have a higher proportion than the median for less vehicles than workers, and lower proportion than the median for dwellings 1/4 mile from a bus stop.

```{r}
selected_tracts <- bus_data %>% 
  filter(propEstHouseholds_LessVehiclesThanWorkers > median(propEstHouseholds_LessVehiclesThanWorkers) &
           propDwell_quarterm_stop < median(propDwell_quarterm_stop))
nrow(selected_tracts)
```

### Statistics: selected tracts

```{r}
totaldwellings_select <- sum(selected_tracts$totalDwellings)
totalhouseholds_select <- sum(selected_tracts$estHouseholds)
totalpopulation_select <- sum(selected_tracts$estPopulation)

cat(paste0(
  "- Most of these tracts have about ", median(selected_tracts$totalRoutes), " bus routes crossing through\n\n",
  
    "According to estimates by the American Community Survey,\n", 
  "  • ", round(sum(selected_tracts$estHouseholds_MorePeopleThanVehicles)/totalhouseholds_select,4)*100, "% of households in these tracts have more people than vehicles.\n",
  "  • ", round(sum(selected_tracts$estHouseholds_1Vehicle_2PlusPeople)/totalhouseholds_select,4)*100, "% of households have one vehicle and more than two people.\n",
  "  • ", round(sum(selected_tracts$estHouseholds_LessVehiclesThanWorkers)/totalhouseholds_select,4)*100, "% of households have less vehicles than workers.\n",
  "  • ", round(sum(selected_tracts$estHouseholds_0Vehicle)/totalhouseholds_select,4)*100, "% of households have 0 vehicles altogether.\n",
  
    "When defining 'home' as every single-family residential building and every individual unit in a multi-family residential building,\n",
  "  • ", round(sum(selected_tracts$quarterm_stop_totalDwellings)/totaldwellings_select,4)*100, "% of homes in these tracts are within walking distance (1/4 mile) of a bus stop.\n",
  "  • ", round(sum(selected_tracts$halfm_stop_totalDwellings)/totaldwellings_select,4)*100, "% of homes are within 1/2 mile of a bus stop.\n",
  "  • ", sum(selected_tracts$quarterm_route_BRT_totalDwellings), " homes are within walking distance (1/4 mile) of a proposed or existing Bus Rapid Transit route/line. This is ", round(sum(selected_tracts$quarterm_route_BRT_totalDwellings)/totaldwellings_select,4)*100, "% of all homes. \n",
"  • ", sum(selected_tracts$halfm_route_BRT_totalDwellings), " homes are within 1/2 mile of a proposed or existing BRT route/line. This is ", round(sum(selected_tracts$halfm_route_BRT_totalDwellings)/totaldwellings_select,4)*100, "% of all homes. \n\n",

  "  • Approximately ", round(sum(selected_tracts$estHouseholds[selected_tracts$is_sunday == 0])/sum(selected_tracts$estHouseholds),4)*100, "% of these households are in a census tract that has no Sunday service, accounting for approximately ", round(sum(selected_tracts$estPopulation[selected_tracts$is_sunday == 0])/sum(selected_tracts$estPopulation),4)*100, "% of Montgomery County's population.\n",
  "  • On Saturdays, most of the selected census tracts have around ", median(selected_tracts$totalSatDepartures), " departures, and ", sum(selected_tracts$totalSatDepartures < 100), " census tracts (", sum(selected_tracts$totalSatDepartures < 100), "/", nrow(selected_tracts), " = ", round(sum(selected_tracts$totalSatDepartures < 100)/nrow(selected_tracts), 3)*100, "% of tracts) have less than 100 departures per day on Saturday, accounting for about ", round(sum(selected_tracts$estPopulation[selected_tracts$totalSatDepartures < 100])/totalpopulation_select, 4)*100, "% of the population."
))
```
